---
editor_options: 
  chunk_output_type: console
---

# Reconstructing Historical Climate

## Load data

Load data prepared for the GAM, and the stack predictor raster.

```{r}
# read data
data = read_csv("data/output/data_for_gam.csv")
stack = terra::rast("data/output/raster_gam_stack.tif")
```

Prepare data covariates and raster.

```{r}
# trim data for distance to coast
data = mutate(
  data,
  coast = coast / 1000,
  coast = coast + rnorm(length(coast), 0.1, sd = 0.01),
  elev = elev + rnorm(length(coast), 1, sd = 0.01),
  lat = lat + rnorm(length(lat), 0, sd = 0.01)
)

# set distance to coast raster in km
values(stack[["coast"]]) = values(stack[["coast"]]) / 1000
```

Pivot longer to separate by variable.

```{r}
# pivot longer
data = pivot_longer(
    data,
    cols = c("ppt", "t_mean", "t_sd"),
    names_to = "climvar"
)
data = dplyr::select(data, !month)
```

## Fit decadal GAMs using best models

```{r}
# assign decade
data = mutate(
  data,
  decade = round(year / 10) * 10
)

# remove recent decades
data = filter(data, year < 1970)

# split by variable, season, and decade
data = nest(
  data, data = !c(climvar, season, decade)
)
```

Prepare model formulas.

```{r}
# temp model has elevation with 3 knots
model_temp = value ~ s(elev, k = 3)

# ppt model has elevation and coast and lat
model_ppt = value ~ s(elev, k = 3) + coast + lat

# assign model formulas
data = mutate(
  data,
  forms = if_else(
    climvar == "ppt", model_ppt, model_temp
  )
)
```

Fit models.

```{r}
# fit models
data = mutate(
  data,
  mod = map2(data, forms, function(df, form) {
    gam(
      formula = form,
      data = df
    )
  })
)
```

## Correct decadal predictions

Get decadal predictions.

```{r}
# save model predictions
model_pred = map(
  data$mod, function(g) {
    predict(stack, g, type = "response") # order matters here
  }
)
```

Add correction layers.

```{r}
# prepare combinations
correction_layers = crossing(
  climvar = c("ppt", "temp"),
  season = c("dry", "rainy")
)

# read file
correction_layers = mutate(
  correction_layers,
  file = "data/spatial/climate_correction/cl_{climvar}_{season}.tif"
)

# join model preds and correction layers
data$pred = model_pred
data = left_join(data, correction_layers)

# read in files and correct
data = mutate(
  data,
  cl = map(file, terra::rast),
  pred_corrected = map2(pred, cl, function(pred, cl) {
    pred = terra::resample(pred, cl)
    pred * cl
  }) 
)
```

## Aggregate decadal predictions

```{r}
# remove excess layers
data = select(
  data, -pred, -cl, -decade
)

time_bin_breaks = c(1850, 1910, 1940, 1970, 2021)

# cut data by breakpoint years
data = mutate(
  data,
  time_bin = cut(year, time_bin_breaks)
)

# aggregate over breakpoint years
data = group_by(data, season, climvar, time_bin) %>%
  summarise(
    mean_pred = list(Reduce(pred_corrected, f = mean))
  )
```

## Save aggregated predictions

```{r}
pmap(data, .f = function(pred_corrected, climvar, season, time_bin) {
  terra::writeRaster(
    pred_corrected,
    file = glue::glue(
      "data/spatial/climate_correction/cl_{climvar}_{season}_{time_bin}.tif"
  )
})
```
