---
editor_options: 
  chunk_output_type: console
---

# Validating semi-Spatial GAMs for climate

## Prepare weather and spatial data

```{r}
# for data
library(tidyverse)
library(sf)

# for gam
library(mgcv)

# for rasters
library(raster)
library(terra)
```

Load data prepared for the GAM, and the stack predictor raster.

```{r}
# read data
data = read_csv("data/output/data_for_gam.csv")
stack = raster::stack("data/output/raster_gam_stack.tif")
```

Prepare data covariates and raster.

```{r}
# trim data for distance to coast
data = mutate(
  data,
  coast = coast / 1000,
  coast = coast + rnorm(length(coast), 0.1, sd = 0.01),
  elev = elev + rnorm(length(coast), 1, sd = 0.01),
  lat = lat + rnorm(length(lat), 0, sd = 0.01)
)

# set distance to coast raster in km
values(stack[["coast"]]) = values(stack[["coast"]]) / 1000
```

Pivot longer to separate by variable.

```{r}
# pivot longer
data = pivot_longer(
    data,
    cols = c("ppt", "t_mean", "t_sd"),
    names_to = "climvar"
)
data = dplyr::select(data, !month)

# select period 1970 -- 2010
data = filter(data, between(year, 1970, 2010))
```

Split the data by variable.

```{r}
data = nest(
  data, data = !c(climvar, season, year_bin)
)
```

## Fit candidate models

Prepare model formulas.

```{r}
# elevation with 3 knots
form_elev_model = value ~ s(elev, k = 3)

# elevation and coast
form_elev_coast = value ~ s(elev, k = 3) + s(coast, k = 3)

# elevation and coast with lat
form_elev_colat = value ~ s(elev, k = 3) + s(coast, lat, k = 5)
```

```{r}
# make combinations
data = crossing(
  data, 
  forms = c(form_elev_model, form_elev_coast, form_elev_colat)
)
```

```{r}
# fit models
data = mutate(
  data,
  mod = map2(data, forms, function(df, form) {
    gam(
      formula = form,
      data = df
    )
  })
)
```

Save model data.

```{r}
save(data, file = "data/output/model_gam_climate.Rds")
```

## Predict over static environment

Get predictions

```{r}
# save model predictions
model_pred = map(
  data$mod, function(g) {
    predict(stack, g, type = "response") # order matters here
  }
)

save(model_pred, file = "data/output/model_pred_climate.Rds")
```

Average predictions over 1970 -- 2010.

```{r}
# assign as list column
data = mutate(
  data,
  pred = model_pred
)

# summarise using a reduce over the list
data_pred = ungroup(data) %>% 
  group_by(season, climvar, forms) %>% 
  summarise(
    mean_pred = list(Reduce(pred, f = mean))
  )

# save averaged predictions
save(
  data_pred,
  file = "data/output/model_pred_average.Rds"
)
```

## Get BIOCLIM data

```{r}

```

